---
title: "R Basics"
author: "Richard Goodman"
output: html_document
---

\

## Knowing where you are 

\

Before you start make sure you have started a new project and know which directory you are working in. 

Check R is working by printing Hello World (use brackets)

```{r, message=FALSE, results='hide'}
print("Hello World")
```
To check you're in the right place use 
```{r, message=FALSE, results='hide'}
getwd()
```
If you're in the wrong place use `setwd()` to change directory.

\

## Assignment

\

In R you can assign values to objects by using `=` or `<-`. This will save your object to the global environment. 

```{r, message=FALSE, results='hide'}

x <- 2 
x = 2 

y <- 5 
y = 5

# stay consistent either = or <-


2+5
x+y
```

To remove that object from the global environment you can use `rm()`

```{r, message=FALSE, results='hide'}

rm(x)
rm(y)

```


Once you have assigned value to objects you can continue to use them in calculations. 

```{r, message=FALSE, results='hide'}

# Make a variable
weight_kg = 55

# print that variable 
print(weight_kg)

# Change weight from kg to grams
weight_kg * 1000
weight_g = weight_kg * 1000

# assign a new value to weight_kg
weight_kg = 57.5 
weight_g = weight_kg * 1000

```



\

## Comparing between values 

\

There is a set of standard operators that we can use to compare two numbers or objects, the results will return the Boolean values `TRUE` or `FALSE` 
* `<` - Greater Than  and `<=` Greater than or equal to
* `>` - Less Than and `<=` less than or equal to
* `==` equal (`=` cannot be used since it assigns values)
* `!=` not equal to

```{r, message=FALSE, results='hide'}

2==2 # Equal 

2!=2 #Inequal 

2 <= 2

```


\

## Functions 

\

Functions execute a defined set of commands and automate a process 

You need an input, the function which has a set of commands and you recieve an output

For example the square root function is: 

```{r, message=FALSE, results='hide'}

sqrt(25)

```


we will use the round function, search in help for how to use 
This will tell you the package, the arguments to use and how to use

```{r, message=FALSE, results='hide'}

?round

# if there are several packages with the same package use:

?base::round

```

Let's use the round function with pi:

```{r, message=FALSE, results='hide'}

# Define pi 
pi = 3.142

# Round to 2 decimal places
round(pi, 2)

# round to 0 decimal places as default (don't define the second argument of digits)
round(pi)
round(pi, 0)

```

When writing the arguments of functions we can use either named or unnamed arguments, generally if it's your first time using the function its best to 

```{r, message=FALSE, results='hide'}

# Example of named argments
round(x = 3.142, digits = 2)

# Example of unnamed argments
round(3.142, 2)

# If you name them you can change the order of arguments
round(digits = 2, x = 3.142) ### this is the good practice 

# Yet this won't work with no  
round(2, 3.142) #wrong!

# Make a variable called result so it can be saved into memory 
result = round(x = 3.142, digits = 2)


```

If you can't find a function in help section use Google 

\

## Data Types 

\

```{r, message=FALSE, results='hide'}

animal = "cat" # example of character, words or strings  
status = TRUE # example of logical 
weight = 50 # example of integar e.g. whole numbers
pi = 3.142 #example of double or dbl

```



If you want to know the data type use the class() function 

```{r, message=FALSE, results='hide'}

vector = c(4,12,7,9)
class(vector) #numeric

```

\

## Data structures

\

Data in R can be structured in different ways such as vectors, factors and dataframes. 

**Vector**: list of items of the same data type e.g. 4,6,9,12 

```{r, message=FALSE, results='hide'}

vector = c(4,12,7,9) # this is one directional list of items, everything in () is a function 

```


**Factor**: categorical data (has to be a character vector), not too popular in tidyverse


```{r, message=FALSE, results='hide'}

# First create a vector of some ordered data using c()
mood = c("unhappy", "awesome", "ok", "awesome", "unhappy") 
# Combine the ordered data together using factor()
factor = factor(mood)

```


**data.frame**: contains tabular data - normally data is loaed into data.frame when reading in a file

```{r, message=FALSE, results='hide'}
# Create three vectors with the necessary information using c()
employee <- c('John Doe','Peter Gynn','Jolie Hope')
salary <- c(21000, 23400, 26800)
startdate <- as.Date(c('2010-11-1','2008-3-25','2007-3-14'))
# Combine the three vectors together as a dataframe using data.frame()
employ_data_frame <- data.frame(employee, salary, startdate)
```


\

## Vectors 

\

To create a vector we must always use the `c()` function, this combines their elements together. 

Vectors can be either numbers or characters, but all the items must be the same datatype. 

An example of integer vector would be 

```{r, message=FALSE, results='hide'}

weight_g = c(50,60,65,82) # num means numeric, both dbl and integar are numeric 

```

An example of character vector would be 
 
```{r, message=FALSE, results='hide'}

animals = c("mouse", "rat", "dog") # chr means character 

```

An example of logical vector would be 
 
```{r, message=FALSE, results='hide'}

logic = c(TRUE, FALSE, FALSE) 

```


We can then investigate the properties of the vector

```{r, message=FALSE, results='hide'}

# Find length of vector with the length() function 
length(animals)
length(weight_g)

# str() fucntion checks the structure of yoru data 
str(weight_g) # a numeric vector of 4 items
str(animals) # a character vector of 3 items 

# Adding extra items to a vector
weight_g = c(weight_g, 90) # add 90 to end of vector 
weight_g = c(30, weight_g) # add 30 to start of vector 

# To add an item to a chosen point within the vector use the append() function
append(x = vector, values = 10, after = 1) 
# x is vector, values are the values to be included, after - after whcih the values will be appended

# Working with heights  
height_mm = c(100,150,99,87)
total_height = sum(height_mm) # sum adds up all items in vector
height_mm = c(220, height_mm)
total_height = sum(height_mm)

```


To look for vector indices we can use `[]`


```{r, message=FALSE, results='hide'}

#get the second item from my vector 
animals[2]
#get the first and second
animals[c(1,2)]
# get multiple items - even repetitions
animals[c(1,1,1,3,2,1,2,2)]

```

R indices start at 1 unlike other programming languages like Python which counts from 0


\

## Factors

\


Factors are categorical variables, they can de split into different types such as nominal, ordinal and binary. 

```{r, message=FALSE, results='hide'}


Nominal = c("turtle", "snail", "butterfly") # unordered descriptions
Ordinal = c("small", "medium", "large") # ordered descriptions
Binary = c("Extinct", "Existing") # only two mutually exclusive outcomes

```


Ordinal variables

```{r, message=FALSE, results='hide'}
# Create a vector containing chr describing mood 
mood = c("unhappy", "awesome", "ok", "awesome", "unhappy") 

```

Convert mood vector to a factor 

```{r, message=FALSE, results='hide'}
factor(mood)
mood = factor(mood)
# factors can only contain pre-defined levels 
# R orders factors in alphabetical order 

#Re-order levels in a factor
factor(mood, levels = c("unhappy", "okay", "awesome"))
```




\

## Dataframes

\

Dataframes are a type of list used to store datasets as tables. The rows are cases and columns are variables, therefore a column can either be a vector or a factor. So we can build a dataframe from the vectors upwards. 

Dataframes can be created using the `data.frame()` function 

```{r, message=FALSE, results='hide'}

# Create three vectors with the necessary information using c()

name <- c('John','Peter','Jolie')
gender <- c('M', 'M', 'F')
age <- c(22, 42, 57)

# Combine the three vectors together as a dataframe using data.frame()

people_data_frame <- data.frame(name, gender, age)


```

To access aspects of the dataframe we can use `$` to subset a dataframe by name 


```{r, message=FALSE, results='hide'}

people_data_frame$name # first column
people_data_frame$gender # second column 
people_data_frame$age # third column 


```

There is more information on working with dataframes in `Importing and Inspecting Data`




\

## Conditional subsetting 

\

We can subset a vector by using a logical vector. 
`TRUE` will select the element with the same index, while `FALSE` will not. 


```{r, message=FALSE, results='hide'}

# which values in weight_g are less than 60

# R asks each item in the vector - "are you less than 60?"
c(TRUE, TRUE, FALSE, FALSE, FALSE, FALSE)
# Apply this to the weight_g vector
weight_g[c(TRUE, TRUE, FALSE, FALSE, FALSE, FALSE)]
# We computed this ourselves as to whne the condition was and wasn't met

# Now let R compute it automatically
weight_g < 60 

```


Now to get the values as an answer, without the `[]` I will not get the values 
Automatic way to get results from queries 

```{r, message=FALSE, results='hide'}
weight_g[weight_g < 60] # greater than 60 
```



Other ways to subset conditonally include

```{r, message=FALSE, results='hide'}
weight_g[weight_g <= 60] # greater than or equal to 60 
weight_g[weight_g == 60] # equal to 60
weight_g[weight_g > 60] # less than 60
weight_g[weight_g >= 60] # less than or equal to 60
```


As well as `TRUE` and `FALSE` we can use other logical expression such as `&`, `|` and `&in%`.  


**using AND `&`**

Both conditions must be true to allow TRUE 
Select values less than 80 and greater than 50 

```{r}

weight_g[weight_g < 80 & weight_g > 50]

```

**using OR `|`**

If one condition is true it will allow TRUE

```{r}
weight_g[weight_g < 80 | weight_g > 50]
```

These are useful for filtering data 

To check if an item is present in a character list 

```{r, message=FALSE, results='hide'}
animals[animals == "rat"]
animals[animals == "cat"] # character(0) means no cat found in vector
```



**using `%in%`**

The `%in%` in operator can be used for character data only

This can be used to look for 100 significnat genes in a type II diabetes gene list of 100 genes


```{r, message=FALSE, results='hide'}
animals %in% c("rat", "dog") # see if two items are present 
animals[animals %in% c("rat", "dog")]
```



As an alterantive to the %in% operator we can use the intersect() function 

```{r, message=FALSE, results='hide'}

intersect(animals, c("rat"))
intersect(animals, c("rat", "dog")) 

```

       

\

## Dealing with missing Data ####

\

```{r, message=FALSE, results='hide'}

heights = c(3, 5, 8, 12, 6)

# introducing missing values
heights = c(3, 5, 8, 12, 6, NA)

```



```{r, message=FALSE, results='hide'}

# calculate the mean (average of heights)
mean(heights) #NA is returned
mean(heights, na.rm = TRUE) # to remove missing values use na.rm 
min(heights, na.rm = TRUE)

```




We can remove missing values from the vector in several ways.


We can use `na.omit`

```{r, message=FALSE, results='hide'}
na.omit(heights) #removes all NAs in object
heights_no_na = na.omit(heights)
```


We can use `is.na()` and the `!` operator

```{r, message=FALSE, results='hide'}
is.na(heights) # returns logical 
!is.na(heights) # to negate the is.na function use ! - so find all things that aren't missing
heights[] #open up heights
heights[!is.na(heights)] # find all values in heights which are not NAs using the negating ! operation
heights_no_na = heights[!is.na(heights)]
```



We can also use `complete.cases`

```{r, message=FALSE, results='hide'}
heights[complete.cases(heights)]
heights_no_na = heights[complete.cases(heights)]
```




